---
title: 'Outlier Detection Using Cluster Analysis'
author: 'Zeynel Cebeci'
#date: '`r format(Sys.Date(), "%B %d, %Y")`'
date: '2017-12-13'
output:
  prettydoc::html_pretty:
    theme: tactile
    highlight: github
    df_print: paged
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Outlier Detection Using Cluster Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
<style>
body {
text-align: justify}
</style>

#Preparing for the analysis

## Install and load the package odetector
This vignette is designed to be used with the package '`odetector`'. You can download the recent version of the package from CRAN with the following command:

```{r, eval=FALSE, message=FALSE, warning=FALSE}
install.packages("odetector")
```

If you have already installed '`odetector`', you can load it into R working environment by using the following command:

```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(odetector)
```

## Load the data set 
We demonstrate outlier detection with '`odetector`' on a synthetic data set consisting of the three features (`p1`, `p2` and `p3`) of four clusters. This three-dimensional data set was created by using the R Package '`MixSim`' (Melnykov et al, 2013). The data set consists of a total of 130 data objects, 30 in each cluster in addition to 10 samples as the outliers at the bottom. 

```{r echo=TRUE, message=FALSE, warning=FALSE, cols.print=5, rows.print=10}
data(x3p4c)
as.data.frame(x3p4c)
x <- x3p4c[,-4]
```
The following command plots the data set by the clusters. The black marked objects in the plot are the outliers. 

```{r fig.width=7, fig.height=6}
pairs(x, col=x3p4c[,4]+1)
```

# Possibilistic Fuzzy C-Means Cluster Analysis
The outlier detection algorithm in the package '`odetector`' uses the typicality degrees which are produced by a possibilistic clustering algorithm such as PCM, FPCM, PFCM or UPFC. Here, we perform the outlier detection process on the results from UPFC algorithm (Wu et al, 2010) implemented in the package '`ppclust`'. For the details see the manual of the R package '`ppclust`' (Cebeci et al, 2017). In order to run UPFC, load the '`ppclust`' into working environment.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
library(ppclust)
```
Since the data set '`x3p4c` has four clusters, we run UPFC for four clusters and display the clustering results with following commands:

```{r echo=TRUE, message=FALSE, warning=FALSE}
res.upfc <- upfc(x, centers=4)
as.data.frame(res.upfc$t)
```

# Outlier Detection

The outlier detection algorithm uses the object of `ppclust` class which is returned by possibilistic and fuzzy clustering algorithm as shown in the previous section. Outlier detection is started with the defaults of the function `detect.outliers` as follows:

```{r echo=TRUE, message=FALSE, warning=FALSE}
res.out <- detect.outliers(res.upfc)
```
In order to change the threshold typicality, `tt` and `rst` arguments can be started with different threshold values. In the following command  `tt` is set to 0.05 for the Approach 1 and `rst` is set to 0.4 for the Approach 2. See the package manual for the details about these arguments.

```{r echo=TRUE, message=FALSE, warning=FALSE}
res.out <- detect.outliers(res.upfc, tt=0.05, rst=0.4)
``` 

## Print the outliers
The result of the `detect.outliers` is an object of `outliers` class. The components of this class can be displayed individually. For example, while the first command displays the result detected with the Approach 1, the second one displays the outliers detected with Approach 2:

```{r echo=TRUE, message=FALSE, warning=FALSE}
res.out$outliers1
res.out$outliers2
```
The full list of outliers obtained with all of the approaches can be displayed together with the function `print.outliers` as follows:

```{r echo=TRUE, message=FALSE, warning=FALSE}
print(res.out)
```

## Summarize the outliers

The function `summary.outliers` is used to summarize the results of outlier detection analysis as follows:

```{r echo=TRUE, message=FALSE, warning=FALSE}
summary(res.out)
```

## Visualiation of the outliers
There are many ways of visual representation of the results of outlier detection analysis. A traditional way is to plot the results by using the functions `plot.outliers` and `pairs.outliers`.

### Plot the outliers
The function `plot.outliers` can be used as follows:

```{r fig.width=7, fig.height=6}
plot(res.out)
```

### Pairwise Scatter Plots
In order to display the outliers by the pairs of features, use the function `pairs.outliers` as follows:

```{r fig.width=7, fig.height=6}
pairs(res.out)
```

## Remove the outliers
For furher analysis of data, the outliers can be removed from the original data set by using the function `remove.outliers` as follows:

```{r echo=TRUE, message=FALSE, warning=FALSE}
Xr <- remove.outliers(res.out, sc=FALSE)
```
In the above command, the option `sc` is set to `TRUE` if the data objects  in small clusters are desired to be treated as collective outliers. Compare the following figure to the figure which has been plotted for the original data in the second section.

```{r fig.width=7, fig.height=6}
pairs(Xr, col=x3p4c[,4]+1)
```

# Using different threshold values 

As default, the outlier detection algorithm uses `tt`, threshold typicality level of 0.05. While the much more outliers is expected with the higher level of this argument, the lower values can be resulted with the less number of outliers. In the following command `tt` has been set to 0.1 instead of the default value of 0.05.

```{r echo=TRUE, message=FALSE, warning=FALSE}
res.out <- detect.outliers(res.upfc, tt=0.1)
```
As seen below, the data object `66` is also evaluated as the outlier with the setting of `tt=0.1`. For the details, see the package manual of '`odetector`'.


```{r echo=TRUE, message=FALSE, warning=FALSE}
res.out$outliers1
```

#References{-}

Wu, X., Wu, B., Sun, J. & Fu, H. (2010). Unsupervised possibilistic fuzzy clustering. *J. of Information & Computational Sci.*, 7 (5): 1075-1080.

Melnykov, V., Chen,W-C. & Maitra, R. (2013). MixSim: An R package for simulating data to study performance of clustering algorithms. *J.  of Statistical Software*, 51(12):1-25.

Cebeci, Z., Yildiz, F., Kavlak, A.T., Cebeci, C. & Onder, H. (2017). ppclust: Probabilistic and Possibilistic Cluster Analysis. R package
  version 0.1.0, URL https://CRAN.R-project.org/package=ppclust
